---
title: "block analysis"
author: "Kai McNamee"
date: "12/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidycensus)
library(leaflet)
library(janitor)
library(sf)
library(ggplot2)
library(geosphere)
library(rstanarm)
library(tidymodels)
library(RColorBrewer)
library(readxl)
library(gtsummary)
library(broom.mixed)
library(gt)
library(ggridges)
```

```{r clean cache=true}

# Read in geospatial data for MBTA lines. st_transform(crs = 4326) stores
# shapefile data in longitude and latitude date (GeoJSON).

mbta <- st_read(dsn = "data/mbta_rapid_transit", 
                      layer = "MBTA_ARC") %>% 
  st_transform(4326) %>%
  clean_names() %>% 
  mutate(line = str_to_title(line)) %>%
  mutate(route = paste(line, "Line,", route))

# Read in geospatial data for MBTA stations. 

mbta_nodes <- st_read(dsn = "data/mbta_rapid_transit/",
                 layer = "MBTA_NODE")%>% 
  st_transform(4326) %>%
  clean_names()

# Extract the longitude and lattitude data from the geometry column in the
# mbta_nodes data. Having the coordinates in 2 columns will be necessary for
# calculations in the next chunk. extract() was giving me issues while piped
# with the rest of the data cleaning, so it has to be executed in its own pipe.

node_extract <- extract(mbta_nodes, geometry, 
                      into = c("lon", "lat"), "\\((.*),(.*)\\)", conv = T)

# Combine the extracted lon-lat data into an object containing station names.
# Add an id column to make it possible to recode the data in later
# transformations.

node_coord <- tibble(station = node_extract$station) %>% 
  mutate(lon = node_extract$lon,
         lat = node_extract$lat,
         id = row_number())

# Read in ACS data. Define reacevars object to pass into the get_acs() call.
# Within get_acs(), specify block group as the unit of analysis, MA as the
# state, the three counties served by the MBTA as the counties, and summary_var
# as the population.

racevars <- c(White = "B02001_002",
              Black = "B02001_003",
              Asian = "B02001_005",
              Hispanic = "B03003_003")

blocks <- get_acs(geography = "block group",
                  variables = racevars, 
                  year = 2018,
                  state = "025",
                  county = c("017", "021", "025"),
                  geometry = TRUE,
                  summary_var = "B02001_001") %>% 
  clean_names() %>% 
  mutate(percent = 100 * estimate / summary_est) %>% 
  st_as_sf() %>% 
  st_transform(4326) %>% 
  arrange(geoid)

# Use st_centroid to calculate the geographic centroids of each census block
# group. Note: the tract analysis uses coordinates for each tract's
# population-weighted center, but I couldn't find equivalent data for block
# groups.

block_centroids <- blocks %>% 
  st_centroid() %>% 
  select(geoid, geometry)

# Extract the coordinate values from the geometry column of block_centroids.

centroid_coord <- extract(block_centroids, geometry, 
                      into = c("lon", "lat"), "\\((.*),(.*)\\)", conv = T) %>% 
  select(geoid, lon, lat)
```

```{r}
# Use pointDistance in the raster package to calculate the distances between
# tract centroids and MBTA nodes. Note: don't load raster package because it
# will break something else. Call raster function with raster::pointDistance.
# The result is a matrix containing the distances between every combination of
# tract centroids and MBTA stations.

distance_matrix <- raster::pointDistance(centroid_coord[c("lon", "lat")],
                           node_coord[c("lon", "lat")], lonlat = TRUE)

# Extract the closest station to each centroid and store it in the pairs object.
# Store the minimum distance to the station for each centroid in the distance
# object. More info: # https://stackoverflow.com/questions/47663254/spatial-nearest-neighbor-assignment-in-r/47664622#47664622

pairs <- apply(distance_matrix, 1, which.min)
distance <- apply(distance_matrix, 1, min)

# Combine distance data with county info. Join node_coord to convert closest
# station id's into closest station names. The final combined object includes
# census tract shape data, demographic data, distance to the nearest T station,
# and name of the nearest T station.

combined <- blocks %>% 
  mutate(closest_node = pairs,
         distance = distance) %>% 
  inner_join(node_coord, by = c("closest_node" = "id")) %>% 
  select(-lon, -lat) %>% 
  mutate(name = str_remove(name, ", Massachusetts"))

# Save the combined data frame as an RDS file for Shiny. 
# saveRDS(combined, "data.RDS")
```

```{r model-data}

# Create a new variable that indicates which demographic group is most prevalent
# in a given census tract. First, drop the geometry data so the data can be
# transformed; group the data by tract, arrange by percent, and slice the top
# row for each group. The result is a new string column that can be joined with
# the model data.

race_plurality <- combined %>% 
  st_drop_geometry() %>%
  group_by(geoid) %>% 
  arrange(desc(percent), .by_group = TRUE) %>% 
  slice(1) %>% 
  ungroup() %>% 
  select(geoid, variable) %>%
  mutate(plurality = variable) %>% 
  select(-variable)

# Clean the combined data. Drop the geometry data so the data frame can be used
# with tidymodels workflow. Filter to include tracts that are a limited distance
# from the nearest T stop. Join the reace_plurality data frame.

model_data <- combined %>% 
  st_drop_geometry() %>%
  select(geoid, variable, percent, distance) %>% 
  pivot_wider(names_from = variable, values_from = percent) %>% 
  clean_names() %>% 
  inner_join(race_plurality, by = "geoid") %>% 
  drop_na() %>% 
  
  # After cross validation (in the model-choice chunk in the tract analysis), I
  # found that there were not enough tracts with mostly Asian residents to
  # meaningfully analyze the effect of plurality = Asian in the model. Filter
  # model_data to drop rows where plurality = Asian to keep the analysis
  # consistent.
  
  filter(distance <= 1609, plurality != "Asian") 
```

```{r fit-obj}
set.seed(10)

fit <- stan_glm(data = model_data, formula = distance ~ white + black + 
           hispanic + plurality - 1,
           family = "gaussian",
           refresh = 0)
```


```{r predictions}
# Define the covariates used for running the predictions. I store 4 rows in the
# new_obs object, each representing tracts by majority racial/ethnic group. For
# each row, I use the average population proportions present in that case. For
# example, row 3 represents majority Asian tracts, and the population
# proportions for row three are the average proportions for majority Asian
# tracts.

new_obs <- model_data %>% 
  group_by(plurality) %>% 
  summarize(white = mean(white),
            black = mean(black),
            asian = mean(asian),
            hispanic = mean(hispanic), .groups = "drop") %>% 
  ungroup()

# Use the fit object and new_obs object to create draws from a posterior
# probability distribution. The pp object stores predictions for individual
# tracts based on the input new_obs. The columns represent the four levels of
# the plurality variable -- white, Black, Asian, Hispanic. The rows represent
# predicted distances for each case with average demographic proportions.

pp <- posterior_predict(fit, newdata = new_obs) %>% 
  as_tibble() %>% 
  mutate_all(as.numeric)

# Rename the columns for clarity,
  
colnames(pp) <- new_obs$plurality

# Create a histogram with the pp object to visualize the posterior probability
# distribution of predicted distances for majority white, Black, Asian, and
# Hispanic tracts, given average demographics in each case. Clean up the labels
# and theme elements.

pp_graph <- pp %>%
  pivot_longer(cols = new_obs$plurality, 
               names_to = "plurality", values_to = "distance") %>% 
  ggplot(aes(x = distance, y = plurality, fill = plurality)) +
    geom_density_ridges(stat = "binline", 
                        bins = 100, 
                        scale = 3,
                        alpha = 0.6,
                        size = 0) +
    scale_fill_brewer(palette = "Dark2",
                      name = "Tract majority",
                      labels = c("Black", "Hispanic", "White")) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(y = "",
         x = "Distance to nearest MBTA station (meters)",
         title = "Posterior Predictive Distribution",
         subtitle = "Predicted distance by census blocks' majority racial/ethnic group",
         caption = "\n*Tracts up to 3218 meters (1 mile) from nearest station")

# Create another posterior distribution to predict the expected outcome for all
# census tracts in the analysis. Use the fit and new_obs objects to generate
# 4000 distance predictions for majority white, Black, Asian, and Hispanic
# tracts, given average demographics in each case. Clean up the labels and theme
# elements.

pe <- posterior_epred(fit, newdata = new_obs) %>% 
  as_tibble() %>% 
  mutate_all(as.numeric)

# Rename the columns for clarity.

colnames(pe) <- new_obs$plurality

# Create a histogram with the pe object to visualize the posterior probability
# distribution of expected distances for majority white, Black, Asian, and
# Hispanic tracts, given average demographics in each case. Clean up the labels
# and theme elements.

pe_graph <- pe %>% 
  pivot_longer(cols = new_obs$plurality, 
               names_to = "plurality", values_to = "distance") %>% 
  ggplot(aes(x = distance, fill = plurality)) +
    geom_histogram(aes(y = after_stat(count/sum(count))),
                   bins = 100,
                   position = "identity",
                   alpha = 0.6) +
    scale_fill_brewer(palette = "Dark2", 
                      name = "Tract majority") +
    scale_y_continuous(labels = scales::percent_format()) +
    theme_minimal() +
    labs(x = "Distance to nearest MBTA station (meters)",
         y = "Probability",
         title = "Posterior Probability Distribution",
         subtitle = "Expected distance by census blocks' majority racial/ethnic group",
         caption = "\n*Tracts up to 1609 meters (1 mile) from nearest station")

# Create a table to store the results of the regression. The table stores the
# coeeficients for each predictor in the model formula.

tbl <- fit %>% 
  tbl_regression(label = list(white ~ "% White", 
                              black ~ "% Black", 
                              hispanic ~ "% Hispanic",
                              plurality ~ "Majority group")) %>% 
  italicize_levels() %>% 
  bold_labels() %>% 
  as_gt() %>% 
  tab_header("Regression of demographics and distance", 
             subtitle = "Effect of demographics on MBTA accessibility") %>% 
  tab_source_note("*Blocks up to 1609 meters (1 mile) from nearest station")

pp_graph
pe_graph
tbl

# Use the following code to save the graphics as images for the Shiny app.
# ggsave("pe_block_1.png",
#        plot = pe_graph,
#        device = "png",
#        scale = 1,
#        width = 7,
#        height = 5,
#        units = "in",
#        dpi = 300,
#        path = "app/www")
```


```{r}
block_map <- blocks_sf

block_white <- block_map %>% 
  filter(variable == "White")

block_black <- block_map %>%
  filter(variable == "Black")

block_asian <- block_map %>%
  filter(variable == "Asian")

block_hispanic <- block_map %>%
  filter(variable == "Hispanic")

# Define palette and icons for the map.

pal <- colorNumeric(palette = "plasma", domain = block_map$percent, n = 100)
pal2 <- colorNumeric(palette = "plasma", domain = block_map$rank, n = 100)

mbtaIcon <- makeIcon("images/mbta-icon.png", iconWidth = 8, iconHeight = 8)

# Create a Leaflet map. Add map panes to order the layers, and add tiles (map
# background).

leaflet() %>% 
  addProviderTiles(provider = "CartoDB.Positron") %>%
  
  # Add a new polygon layer for each demographic layer, as defined in lines
  # 167-181. Define the popup labels for each tract. Note: the Shiny version of
  # this code uses a seperate plot output instead of the native Leaflet popups.
  
  addPolygons(data = block_white,
              highlightOptions = highlightOptions(fillOpacity = 0.3),
              stroke = FALSE,
              smoothFactor = 0,
              fillOpacity = 0.7,
              color = ~ pal(percent),
              group = "White") %>% 


